The Design and Implementation of the Synchronous Language \CEU

### 2017-01-08 paper.tex
(>) typo
(>) images
(>) apply minor suggestions

Despite its success and influence, Esterel has a complex semantics that
requires careful static analysis to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems%
%These problems have an extensive coverage in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
A complex semantics not only challenges the analysis and compilation of 
programs, but also affects the programmer's understanding about the code, who,
ultimately, has to solve the errors when facing corner cases.
%
Furthermore, Esterel's semantics is non-deterministic for intra-reaction
statements, which prevents threads from interacting with stateful system calls
safely and makes shared-memory concurrency not as straightforward as reading
and writing to shared variables.
%
%For instance, although Esterel supports orthogonal abortion of 
%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
%release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in some fundamental 
semantic aspects.
%
\CEU proposes a semantics for soft real-time systems with fine-grained control
for intra-reaction execution which is amenable to concurrency checks that
improve safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static concurrency checks to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.checks}).
%
\item Safe integration with \C that enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of the proposed semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

The semantics of \CEU implies an implementation with a runtime stack and a
sequential scheduler.
Unlike Esterel's semantics, this prevents programs to be compiled directly and
efficiently to hardware~\cite{esterel.constructive}.
It also makes difficult to statically determine worst-case reaction times
towards hard real-time systems~\cite{esterel.fixed}.

@@

\CEU is a synchronous reactive language inspired by Esterel in which programs 
advance in a sequence of discrete reactions to external events.
%
%Regarding the similarities with Esterel,
%\CEU has a strong imperative flavor, with explicit control flow through 
%sequences, loops, and parallel compositions.
%
Like Esterel, \CEU is designed for control-intensive applications, supporting 
concurrent lines of execution, known as \emph{trails}, and broadcast 
communication through events.
%
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
An \code{await} is the only statement that halts a running reaction and allows 
a program to advance in this discrete notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with (core) Esterel and synchronous languages 
in general~\cite{esterel.preemption}:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis, and cannot be directly implemented.

Figure~\ref{lst.abro} illustrates the syntactic similarities between the 
languages, showing side-by-side the implementations in Esterel \ax and \CEU \bx 
for the following control specification:
%
\emph{``Emit an output O as soon as inputs A and B occur.
Reset this behavior each time input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 5--10, in both 
implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses an 
\code{abort-when} statement (ln. 4--11) which, in this case, serves the same 
purpose as \CEU's \code{par/or} (ln. 4--13):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

Esterel defines time as a discrete sequence of logical unit instants or 
``ticks''.
At each tick, the program reacts to an arbitrary number of simultaneous input 
events from the environment.
In contrast, \CEU defines time as a discrete sequence of reactions to
unique input events.
At each input event, which constitutes a logical unit of time, the program 
reacts exclusively to it.
The event-triggered execution of a program in \CEU is as 
follows~\cite{ceu.sensys13}:
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail (but 
      parallel constructs may create new trails).
\item Active trails execute until they await or terminate, one after the other.
      This step is named a \emph{reaction chain}, and always runs in bounded 
      time.
\item The program goes idle and the environment takes control.
\item On the occurrence of a new external input event, the environment awakes 
      \emph{all} trails awaiting that event.
      It then goes to step 2.
\end{enumerate}
%
A program must react to an event completely before handling the next 
one.
%
Based on the synchronous hypothesis, a program takes a negligible time on step 
2 and is always idle on step 3.
In practice, if a new external input event occurs while a reaction chain is 
running, it is enqueued to occur in a subsequent reaction.

\begin{figure}[!htb]
\centering
\includegraphics[width=\columnwidth]{tick}
\caption{The discrete notions of time in Esterel and \CEU.
\label{fig.ticks}
}
\end{figure}

Figure~\ref{fig.ticks} compares the discrete notions of time in two variations 
of Esterel and in \CEU.
The box \code{Real World} assumes an external observer with an absolute
reference clock which timestamps event occurrences over a
continuous timeline~\cite[Chapter~3]{es.time} (e.g., event \code{C} occurs at
$17ms521us$).
The other boxes show how the same occurring events fit differently in each 
logical notion of time.
%
\begin{itemize}
\item \code{[Box-1]}:
\emph{Esterel with fixed-length ticks}~\cite{esterel.fixed}.
%
(Usually referred as \emph{sample-driven execution}~\cite{rp.twelve}.)
%
We assume a reaction \code{R(boot)} at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes in the next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1} and invades 
\code{tick-2}, causing a \emph{timing violation}~\cite{esterel.fixed}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur during the idle \code{tick-3}.
However, only one occurrence of \code{E} can be considered in \code{R(D+E)}.
%
\item \code{[Box-2]}:
\emph{Esterel with variable-length ticks}~\cite{esterel.variable}.
This approach avoids the timing violation for \code{R(A)} and also results in
smaller idle periods because it adjusts the tick lengths to match the CPU times 
for the reactions.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
react alone as \code{R(D)} at \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered at \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{[Box-3]}:
\emph{\CEU with unique and queue-based input events}.
We also assume a reaction \code{R(boot)} before any input.
%
Because the occurrence of event \code{A} is unique during \code{R(boot)}, the 
behavior in \CEU is similar to \code{Box-2} for its first two reactions 
(\code{tick-0} and \code{tick-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} as 
simultaneous, and handles each in subsequent reactions \code{R(B)} and 
\code{R(C)}.
%
We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
\CEU to be roughly the same.
This way, the first idle periods in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU reacts to the two instances of \code{E} independently, which are
handled in sequence.
%
%Note that, by definition, each reaction corresponds to an event occurrence of 
%the same length.
\end{itemize}

Sample-driven execution maps directly to hardware implementations and
involve no runtime queues.
%
However, considering the soft real-time application domain of \CEU, we
decided for the unique and queue-based semantics in \CEU for the reasons
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is implementation dependent:}
Tick lengths are not part of the program specification.
For instance, if we log the \code{Real World} box timeline and reproduce it in
implementations with different tick lengths, the behaviors might diverge.
%
\item \textbf{Events are never absolutely simultaneous:}
We consider that the notion of simultaneity should be defined explicitly for
each use case (to be discussed in Section~\ref{sec.ceu.timers}).
%In Section~\ref{sec.ceu.timers}, after introducing internal events and 
%first-class timers, we show an example to illustrate how to detect simultaneous 
%button clicks in \CEU.
In tick-based approaches, simultaneity depends on the length of discrete ticks.
For instance, in \code{box-1} and \code{box-2} of Figure~\ref{fig.ticks}, 
events \code{B} and \code{C} are simultaneous, even though \code{A} and
\code{B} ``physically'' happen much closer to one another.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to exclusive events are atomic and never overlap.
Automatic mutual exclusion simplifies reasoning about concurrency and is a 
prerequisite for the concurrency checks to be discussed in 
Section~\ref{sec.ceu.checks}.
\end{itemize}

@@

In Esterel, the behavior of internal and external signals is equivalent.
%In particular, programs can emit external and internal signals simultaneously, 
%with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
%In \CEU, subsequent external events coming from the environment define a 
%timeline, which programs cannot interfere by emitting new external events.
%
%To cope with deterministic intra-reaction communication, \CEU supports 
%internal events which programs can \code{emit} and \code{await}.
In \CEU, in contrast with queue-based external events, internal events follow a 
stack-based execution policy similar to subroutine calls in typical programming 
languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel \ax and \CEU \bx.
%
In Esterel, when \code{A} occurs, the program emits \code{B} (ln.  4--5) and 
both events become active, resulting in the invocation of \code{f()} and 
\code{g()} in no particular order (ln. 6,9).
%
In \CEU, when \code{A} occurs, the program behaves as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes, broadcasts \code{b}, and pauses (ln. 4--5).
\item 2nd trail awakes, calls \code{\_g()}, and terminates (ln. 8--9).
      \emph{(No other trails awake to \code{b}.)}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates (ln. 5--6).
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

Internal events provide fine-grained execution control and can express a 
limited form of subroutines, as depicted in Figure~\ref{lst.sub}.
The ``subroutine'' \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed by
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine through an \code{emit 
inc} (ln. 10) in reaction to some code (ln. 9).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence, the calling trail resumes and passes the assertion 
test (ln. 10--11).

@@

The example in Figure~\ref{lst.await} illustrates delayed awaits, which 
prevents infinite execution in the loop.
Both sides of the \code{par/or} have an \code{await} statement (ln. 4,7), which 
characterizes the enclosing \code{loop} as non instantaneous (ln 2--9).
However, if the \code{emit e} (ln. 6) could awake the \code{await e} 
instantaneously (ln. 4), the \code{par/or} would terminate and restart the 
\code{loop} also instantaneously, resulting in infinite execution.
%
\begin{comment}
\begin{figure}
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits some period
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute immediately and then periodically
        await e;  // await after
    end
end
\end{lstlisting}
%\end{minipage}
%
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[numbers=left,xleftmargin=3em]
%event void e;
%var bool should_execute = false;
%par do
%    loop do
%        <...>         // code that awaits
%        if <...> then // some conditon
%            should_execute = true;
%            emit e;
%        end
%    end
%with
%    loop do
%        if should_execute then
%            <...>     // code to execute
%        end
%        await e;
%    end
%end
%\end{lstlisting}
%\end{minipage}
%
\caption{ An example that circumvents the \emph{delayed await} by post-fixing 
the \code{await} inside the \code{loop}.
% (in the left), and by copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}
\end{comment}
%
In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by placing the code to execute before the \code{await}.
%
\begin{comment}
For instance, sometimes we need to execute a block of code immediately, and 
then periodically from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
\end{comment}
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for complex analysis to deal with schizophrenic statements.

@@

To mitigate this threat, \CEU performs concurrency checks at compile time to 
detect conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment 
cannot read or write to that variable, nor dereference a pointer of that 
variable type.
%An analogous policy is applied for pointers \emph{vs} variables and pointers 
%\emph{vs} pointers.
%
Concurrency in \CEU is characterized when two or more trail segments in 
parallel react to the same input event.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
Considering the examples in Figure~\ref{lst.shared}:
%
\begin{itemize}
\item The assignments to \code{x} (\ax:2,5)
      \textbf{cannot} be concurrent because they are \textbf{not} in parallel trails.
\item The assignments to \code{x} (\ax:5,8)
      \textbf{cannot} be concurrent because they \textbf{cannot} execute during 
      the same reaction.
\item The assignments to \code{y} (\bx:5,8)
      \textbf{can} be concurrent because they are in parallel trails and 
      \textbf{can} execute during the same reaction.
\end{itemize}
%
The detection algorithm, which is depicted in Section~\ref{sec.impl.checks},
inspects all possible \code{await} statements that precede a variable access
and keeps a list with all corresponding awaking events.
Then, it checks all accesses in parallel trails to see if they share an awaking 
event.
If it is the case, the compiler warns about the suspicious accesses.

@@

From our experience, however, we find that programs often need non-commutative 
concurrent calls.
%
This is the case for logging (e.g., calls to \code{\_printf} in trails in 
parallel) and for redrawing objects in the screen.
%
Figure~\ref{lst.redraw}.a shows an abstract code to animate and redraw the 
objects \code{background} and \code{foreground} in trails in parallel.
In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
conflicting pixels, which makes the calls non commutative and prevents the code 
to compile.
%
However, in this case we want to rely on lexical order to always redraw the 
\code{background} object before the \code{foreground} object.
%
Therefore, in Figure~\ref{lst.redraw}.b, we redefine \code{\_redraw} as
\code{\_redraw\_non\_commutative} (ln. 2), communicating its effect explicitly,
and annotate it as safe (ln. 4--5) to make the code compile successfully.

@@

The compilation process of programs in \CEU is composed of three main phases:
the \emph{parsing phase} converts the source code in \CEU to an abstract syntax 
tree (\emph{AST});
the \emph{concurrency checks phase} detects inconsistencies in programs, such
as unbounded loops and suspicious concurrent statements;
the \emph{code generation phase} converts the \emph{AST} to standard \C code 
and augments it with platform-dependent functionality (e.g., system calls) and
the runtime of \CEU, compiling everything with \emph{gcc} to generate the final 
binary.
%
An alternative implementation generates bytecode for an embedded virtual
machine, which can disseminate and install programs dynamically.

@@

With all sets calculated, we take all pairs of nodes that perform side effects 
and are in parallel branches, and compare their sets $I$ for intersections.
For each pair, if the intersection is not empty, we mark both nodes as 
suspicious.

>>>>

Despite its success and influence, Esterel has a complex semantics that
requires careful static analysis to detect and reject programs with 
\emph{causality} and \emph{schizophrenia} problems%
%These problems have an extensive coverage in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
A complex semantics not only challenges the analysis and compilation of 
programs, but also affects the programmer's understanding about the code, who,
ultimately, has to solve the errors arising from corner cases.
%
Furthermore, Esterel's semantics is non-deterministic for intra-reaction
statements, which prevents threads from interacting with stateful system calls
safely and makes shared-memory concurrency not as straightforward as reading
and writing to shared variables.
%
%For instance, although Esterel supports orthogonal abortion of 
%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
%release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in some fundamental 
semantic aspects.
%
\CEU proposes a semantics for soft real-time systems with fine-grained control
for intra-reaction execution which is amenable to concurrency checks that
improve safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static concurrency checks to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.checks}).
%
\item Safe integration with \C that enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers,%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
} showing a practical aspect of the proposed semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

The semantics of \CEU implies an implementation with a runtime stack and a
sequential scheduler.
Unlike Esterel's semantics, this precludes programs being compiled directly and
efficiently to hardware~\cite{esterel.constructive}.
It also makes difficult to statically determine worst-case reaction times for
hard real-time systems~\cite{esterel.fixed}.

@@

\CEU is a synchronous reactive language inspired by Esterel in which programs 
advance in a sequence of discrete reactions to external events.
%
%Regarding the similarities with Esterel,
%\CEU has a strong imperative flavor, with explicit control flow through 
%sequences, loops, and parallel compositions.
%
Like Esterel, \CEU is designed for control-intensive applications, supporting 
concurrent lines of execution, known as \emph{trails}, and broadcast 
communication through events.
%
Internal computations within a reaction (e.g., expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
An \code{await} is the only statement that halts a running reaction and allows 
a program to advance in this discrete notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
each possible path~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with (core) Esterel and synchronous languages 
in general~\cite{esterel.preemption}:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis, and cannot be directly implemented.

Figure~\ref{lst.abro} illustrates the syntactic similarities between the 
languages, showing side-by-side the implementations in Esterel \ax and \CEU~\bx 
of the following control specification:
%
\emph{``Emit an output O as soon as inputs A and B occur.
Reset this behavior each time input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 5--10, in both 
implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
analogous.
%
For the second phrase, the reset behavior, the Esterel version uses an 
\code{abort-when} statement (ln. 4--11) which, in this case, serves the same 
purpose as \CEU's \code{par/or} (ln. 4--13):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

Esterel defines time as a discrete sequence of logical unit instants or 
``ticks''.
At each tick, the program reacts to an arbitrary number of simultaneous input 
events from the environment.
In contrast, \CEU defines time as a discrete sequence of reactions to
unique input events.
At each input event, which constitutes a logical unit of time, the program 
reacts exclusively to it.
The event-triggered execution of a program in \CEU is as 
follows~\cite{ceu.sensys13}:
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail (but 
      parallel constructs may create new trails).
\item Active trails execute until they await or terminate, one after another.
      This step is named a \emph{reaction chain}, and always runs in bounded 
      time.
\item The program goes idle and the environment takes control.
\item On the occurrence of a new external input event, the environment awakes 
      \emph{all} trails awaiting that event.
      It then goes to step 2.
\end{enumerate}
%
A program must react to an event completely before handling the next 
one.
%
Based on the synchronous hypothesis, a program takes negligible time on step 
2 and is always idle on step 3.
In practice, if a new external input event occurs while a reaction chain is 
running, it is enqueued to occur in a subsequent reaction.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{tick}
\caption{The discrete notions of time in Esterel and \CEU.
\label{fig.ticks}
}
\end{figure}

Figure~\ref{fig.ticks} compares the discrete notions of time in two variations 
of Esterel and in \CEU.
The box \code{Real World} assumes an external observer with an absolute
reference clock which timestamps event occurrences over a
continuous timeline~\cite[Chapter~3]{es.time} (e.g., event \code{C} occurs at
$17ms521us$).
The other boxes show how the same occurring events fit differently in each 
logical notion of time.
%
\begin{itemize}
\item \code{[Box-1]}:
\emph{Esterel with fixed-length ticks}~\cite{esterel.fixed}.
%
(Usually referred to as \emph{sample-driven execution}~\cite{rp.twelve}.)
%
We assume a reaction \code{R(boot)} at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes in the next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1} and invades 
\code{tick-2}, causing a \emph{timing violation}~\cite{esterel.fixed}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur during the idle \code{tick-3}.
However, only one occurrence of \code{E} can be considered in \code{R(D+E)}.
%
\item \code{[Box-2]}:
\emph{Esterel with variable-length ticks}~\cite{esterel.variable}.
This approach avoids the timing violation for \code{R(A)} and also results in
smaller idle periods because it adjusts the tick lengths to match the CPU times 
for the reactions.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
react alone as \code{R(D)} at \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered at \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{[Box-3]}:
\emph{\CEU with unique and queue-based input events}.
We also assume a reaction \code{R(boot)} before any input.
%
Because the occurrence of event \code{A} is unique during \code{R(boot)}, the 
behavior in \CEU is similar to \code{Box-2} for its first two reactions 
(\code{tick-0} and \code{tick-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} as 
simultaneous, and handles each in subsequent reactions \code{R(B)} and 
\code{R(C)}.
%
We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
\CEU to be roughly the same.
This way, the first idle periods in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU reacts to the two instances of \code{E} independently, which are
handled in sequence.
%
%Note that, by definition, each reaction corresponds to an event occurrence of 
%the same length.
\end{itemize}

Sample-driven execution maps directly to hardware implementations and does not
require runtime queues.
%
However, considering the soft real-time application domain of \CEU, we
decided for the unique and queue-based semantics in \CEU for the reasons
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is implementation dependent:}
Tick lengths are not part of the program specification.
For instance, if we log the \code{Real World} box timeline and reproduce it in
implementations with different tick lengths, the behaviors might differ.
%
\item \textbf{Events are never absolutely simultaneous:}
We consider that the notion of simultaneity should be defined explicitly for
each use case (to be discussed in Section~\ref{sec.ceu.timers}).
%In Section~\ref{sec.ceu.timers}, after introducing internal events and 
%first-class timers, we show an example to illustrate how to detect simultaneous 
%button clicks in \CEU.
In tick-based approaches, simultaneity depends on the length of discrete ticks.
For instance, in \code{box-1} and \code{box-2} of Figure~\ref{fig.ticks}, 
events \code{B} and \code{C} are simultaneous, even though \code{A} and
\code{B} ``physically'' happen much closer to one another.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to exclusive events are atomic and never overlap.
Automatic mutual exclusion simplifies reasoning about concurrency and is a 
prerequisite for the concurrency checks to be discussed in 
Section~\ref{sec.ceu.checks}.
\end{itemize}

@@

In Esterel, the behavior of internal and external signals is equivalent.
%In particular, programs can emit external and internal signals simultaneously, 
%with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
%In \CEU, subsequent external events coming from the environment define a 
%timeline, which programs cannot interfere by emitting new external events.
%
%To cope with deterministic intra-reaction communication, \CEU supports 
%internal events which programs can \code{emit} and \code{await}.
In \CEU, in contrast with queue-based external events, internal events follow a 
stack-based execution policy similar to subroutine calls in typical programming 
languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel \ax and \CEU~\bx.
%
In Esterel, when \code{A} occurs, the program emits \code{B} (ln.  4--5) and 
both events become active, resulting in the invocation of \code{f()} and 
\code{g()} in no particular order (ln. 6,9).
%
In \CEU, when \code{A} occurs, the program behaves as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes, broadcasts \code{b}, and pauses (ln. 4--5).
\item 2nd trail awakes, calls \code{\_g()}, and terminates (ln. 8--9).
      \emph{(No other trails awake to \code{b}.)}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates (ln. 5--6).
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

Internal events provide fine-grained execution control and can express a 
limited form of subroutines, as described in Figure~\ref{lst.sub}.
The ``subroutine'' \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed by
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine through an \code{emit 
inc} (ln. 10) in reaction to some code (ln. 9).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence does the calling trail resumes and passes the assertion 
test (ln. 10--11).

@@

The example in Figure~\ref{lst.await} illustrates delayed awaits, which 
prevents infinite execution in the loop.
Both sides of the \code{par/or} have an \code{await} statement (ln. 4,7), which 
characterizes the enclosing \code{loop} as non instantaneous (ln 2--9).
However, if the \code{emit~e} (ln. 6) could awake the \code{await e} 
instantaneously (ln. 4), the \code{par/or} would terminate and restart the 
\code{loop} also instantaneously, resulting in infinite execution.
%
\begin{comment}
\begin{figure}
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits some period
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute immediately and then periodically
        await e;  // await after
    end
end
\end{lstlisting}
%\end{minipage}
%
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[numbers=left,xleftmargin=3em]
%event void e;
%var bool should_execute = false;
%par do
%    loop do
%        <...>         // code that awaits
%        if <...> then // some conditon
%            should_execute = true;
%            emit e;
%        end
%    end
%with
%    loop do
%        if should_execute then
%            <...>     // code to execute
%        end
%        await e;
%    end
%end
%\end{lstlisting}
%\end{minipage}
%
\caption{ An example that circumvents the \emph{delayed await} by post-fixing 
the \code{await} inside the \code{loop}.
% (in the left), and by copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}
\end{comment}
%
In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by placing the code to execute before the \code{await}.
%
\begin{comment}
For instance, sometimes we need to execute a block of code immediately, and 
then periodically from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
\end{comment}
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for complex analysis to deal with schizophrenic statements.

@@

To mitigate this threat, \CEU performs concurrency checks at compile time to 
detect conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment 
cannot read or write to that variable, nor dereference a pointer of that 
variable type.
%An analogous policy is applied for pointers \emph{vs} variables and pointers 
%\emph{vs} pointers.
%
Concurrency in \CEU is characterized when two or more trail segments in 
parallel react to the same input event.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
Considering the examples in Figure~\ref{lst.shared}:
%
\begin{itemize}
\item The assignments to \code{x} (\ax:2,5)
      \textbf{cannot} be concurrent because they are \textbf{not} in parallel trails.
\item The assignments to \code{x} (\ax:5,8)
      \textbf{cannot} be concurrent because they \textbf{cannot} execute during 
      the same reaction.
\item The assignments to \code{y} (\bx:5,8)
      \textbf{can} be concurrent because they are in parallel trails and 
      \textbf{can} execute during the same reaction.
\end{itemize}
%
The detection algorithm, which is described in Section~\ref{sec.impl.checks},
inspects all possible \code{await} statements that precede a variable access
and keeps a list with all corresponding awaking events.
Then, it checks all accesses in parallel trails to see if they share an awaking 
event.
If it is the case, the compiler warns about the suspicious accesses.

@@

From our experience, however, we find that programs often need non-commutative 
concurrent calls.
%
This is the case for logging (e.g., calls to \code{\_printf} in trails in 
parallel) and for redrawing objects on the screen.
%
Figure~\ref{lst.redraw}.a shows an abstract code to animate and redraw the 
objects \code{background} and \code{foreground} in trails in parallel.
In typical graphical APIs, consecutive calls to \code{\_redraw} overwrite 
conflicting pixels, which makes the calls non-commutative and prevents the code 
from compiling.
%
However, in this case we want to rely on lexical order to always redraw the 
\code{background} object before the \code{foreground} object.
%
Therefore, in Figure~\ref{lst.redraw}.b, we redefine \code{\_redraw} as
\code{\_redraw\_non\_commutative} (ln. 2), communicating its effect explicitly,
and annotate it as safe (ln. 4--5) to make the code compile successfully.

@@

The compilation of \CEU programs consists of three main phases:
the \emph{parsing phase} converts the source code in \CEU to an abstract syntax 
tree (\emph{AST});
the \emph{concurrency checks phase} detects inconsistencies in programs, such
as unbounded loops and suspicious concurrent statements;
the \emph{code generation phase} converts the \emph{AST} to standard \C code 
and augments it with platform-dependent functionality (e.g., system calls) and
the runtime of \CEU, compiling everything with \emph{gcc} to generate the final 
binary.
%
An alternative implementation generates bytecode for an embedded virtual
machine, which can disseminate and install programs dynamically.

@@

With all sets calculated, we take all pairs of nodes that perform side effects 
and are in parallel branches, and compare their sets $I$ for non-empty
intersections.
For each pair, if the intersection is not empty, we mark both nodes as 
suspicious.
